---
title: "CodeClimateCorrelation"
subtitle: "Moving-window climate response surfaces, one site at a time"
format: html
editor: visual
---

This document reproduces the **first analytical section of `CodeMain.qmd`** --
everything up to and including *Moving-window climate correlations* -- for all
six sites that have crossdated cores (AMI, GHE, KAT, LIN, STO, TRO) instead of
for Ghent alone. Nothing that follows that section in `CodeMain.qmd`
(extreme-year selection, drought progression, growth in the extreme years, the
built-up GLMM) is reproduced here.

What changed relative to `CodeMain.qmd`, and why:

- **The `# Only Ghent for now` filter is gone.** Everything up to the
  chronologies now runs once on the full dataset, and the site becomes a
  grouping variable (`Site`, the first three characters of `TreeID`) from the
  chronologies onwards.
- **Chronologies are built per site x species**, and the replication floor is a
  setting (`MIN_REPL`) rather than the hard-coded `BBI_n >= 10`. Ghent has
  69-99 crossdated trees per species; the other five sites have 5-8, so a floor
  of 10 would silently return five empty chronologies. The realised replication
  is printed per site so it is visible what each surface rests on.
- **Latitude and elevation are per site.** `SITE_LAT`/`SITE_ELEV` were two
  scalars; extraterrestrial radiation changes by roughly 10 % per 5 degrees of
  latitude, and these sites span 49.99 N to 63.45 N, so one shared value would
  bias PET badly for Trondheim.
- **The daily climate and the response surfaces are computed in a loop over
  sites**, and both are cached to disk. Re-running the document does not
  recompute a site whose correlation field is already on disk unless
  `RECOMPUTE` is `TRUE`.
- Two small robustness fixes that the single-site version did not need: the
  `Species` spellings in `Trees_Cooltree.xlsx` are normalised before the
  shrinkage join (the sheet contains both `Fraxinus exelsior` and
  `Fraxinus excelsior`, and both `Tillia` and `Tilia`; a typo would join to
  `Sr = NA` and take out `bakker_bai()` with an opaque error), and the ERA5
  readers resolve their own file names and trim incomplete days at either end
  of the record rather than hard-coding `>= 1940-01-02`.

The output is one correlation field per site, saved as
`Output/ClimateCorrelation/<SITE>/ClimCor_<SITE>.rds`, plus a combined
`ClimCor_AllSites.rds` for the sections that will follow.

# Setup

```{r setup, include=FALSE}

#### clear global environment ####
remove(list = ls())

#### load necessary (and usefull) packages ####
library("dplR")       # dendrochronology
library("tidyverse")  # data manipulation
library("dplyr")      # data manipulation
library("ggplot2")    # plotting
library("readxl")     # read xlsx files
library("writexl")    # write xlsx files
library("ggpubr")     # arrange ggplots
library("mgcv")       # GAM (generalized additive model)

#### load XCT.Read function ####
source("XCT.Read.R")

```

# Site configuration

```{r sites}

# The six sites that have crossdated cores. `Site` is the first three
# characters of every TreeID at that site and the suffix of the ERA5 file;
# `Region` is the spelling used in Trees_Cooltree.xlsx. The three regions in
# the metadata with no cored trees (Alnarp, Bremen, Potsdam) drop out by
# themselves at the Cored/Crossdate_Status filter and are deliberately absent
# here, so a new site appearing in the tree-ring data raises an error instead
# of quietly running without a latitude.
#
# Lat/Lon are the mean coordinates of that site's trees, and Lat is only a
# FALLBACK: if the ERA5 csv still carries its own `latitude` column, the
# grid-cell latitude from the file is used instead, because that is the
# latitude the climate series actually belongs to.
#
# Elev is approximate. It enters only the clear-sky reference radiation Rso
# and the psychrometric constant, both weakly (FAO-56), so a rough figure is
# fine -- but it is per site rather than one shared number, because Katowice
# sits some 260 m above the other five.
SITES <- tribble(
  ~Site, ~Region,      ~Lat,    ~Lon,     ~Elev,
  "AMI", "Amiens",     49.9878,  2.3582,   35,
  "GHE", "Ghent",      51.0506,  3.7206,   10,
  "KAT", "Katowice",   50.4520, 18.7579,  270,
  "LIN", "Linkoping",  58.4084, 15.5654,   55,
  "STO", "Stockholm",  59.4410, 18.0915,   30,
  "TRO", "Trondheim",  63.4520, 10.5375,   30
)

#### paths ####
dataFolder             <- "Data/TreeRingData/Raw_100dens_30area"
removeDensityRingsPath <- "Data/TreeRingData/rings not usable for density mesurements.xlsx"
metaDataPath           <- "Data/TreeRingData/Trees_Cooltree.xlsx"
climDir                <- "Data/ClimateData"
outRoot                <- "Output/ClimateCorrelation"

dir.create(outRoot, showWarnings = FALSE, recursive = TRUE)

#### chronology settings ####

# Replication floor. CodeMain used 10, which is right for Ghent (69-99
# crossdated trees per species) and empties every other site (5-8 per
# species). Lower it and the surfaces exist but rest on fewer trees; raise it
# and the non-Ghent sites disappear. Whatever it is set to, the realised
# replication is printed below.
MIN_REPL <- 5

# ERA5 starts in 1940, so nothing before it can be correlated with climate.
FIRST_YEAR <- 1940

# A site x species chronology shorter than this cannot reach `MIN_N`
# overlapping years in any window, so it is dropped before the correlations
# rather than producing an all-NA surface.
MIN_YEARS <- 25

#### recompute or reuse ####

# The correlation fields are expensive (see the timing note in the
# correlations chunk). Set to TRUE to force a recompute; FALSE reuses any
# ClimCor_<SITE>.rds already on disk.
RECOMPUTE <- FALSE

#### find each site's ERA5 file ####

# The file naming is resolved rather than assumed, because CodeMain read
# "ERA5-Ghent.csv" and the six-site set is named "ERA5_AMI" etc.
era5_path <- function(site, region, dir = climDir) {

  cand <- file.path(dir, c(
    paste0("ERA5_", site,   ".csv"),
    paste0("ERA5-", site,   ".csv"),
    paste0("ERA5_", region, ".csv"),
    paste0("ERA5-", region, ".csv"),
    paste0("ERA5",  site,   ".csv")
  ))

  hit <- cand[file.exists(cand)]

  if (!length(hit)) {
    stop("No ERA5 file found for site ", site, " (", region, ").\n  Tried:\n    ",
         paste(cand, collapse = "\n    "),
         "\n  Files present in ", dir, ":\n    ",
         paste(list.files(dir), collapse = "\n    "))
  }

  hit[1]
}

# Resolve all six now: a typo should fail in two seconds, not after the
# tree-ring pipeline has run.
SITES <- SITES %>%
  mutate(ClimPath = purrr::map2_chr(Site, Region, era5_path))

SITES

```

# Load and filter data

```{r}

# Load RW and mean ring density
Data <- XCT.read(
  path = dataFolder,
  output = "ringwidth_density",
  densityType = "fraction",
  area = c(0, 1),
  fun = "mean") %>%
  mutate(TreeID = substr(Sample, 1, 8)) %>%
  na.omit()


# remove -A and -B at the end of split Samples
Data <- Data %>%
  mutate(Sample = str_remove(Sample, "-A$")) %>%
  mutate(Sample = str_remove(Sample, "-B$"))


# remove density of rings with non usefull density data, make those NA
Data <- Data %>%
  full_join(read_excel(removeDensityRingsPath), by = c("Sample", "Year")) %>%
  mutate(Density = ifelse(!is.na(`Reason`), NA, Density)) %>%
  select(-c(`Reason`))


# add temparary species column
Data <- Data %>%
  mutate(
    species_temp = substr(Sample, 5, 5),
    species_temp = recode(
      species_temp,
      "Q" = "Quercus",
      "T" = "Tilia",
      "F" = "Fraxinus"
    )
  )


# what came in, per site
Data %>%
  mutate(Site = substr(Sample, 1, 3)) %>%
  group_by(Site, species_temp) %>%
  summarise(n_samples = n_distinct(Sample), n_rings = n(), .groups = "drop") %>%
  pivot_wider(names_from = species_temp, values_from = c(n_samples, n_rings)) %>%
  print(n = 30)


# plot density vs RW, one panel per species
p1 <- ggplot(Data[is.na(Data$Density)==FALSE, ], aes(x = RW, y = Density)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_wrap(~species_temp) +
  labs(x = "Ring Width (mm)", y = "Mean Ring Density (kg/m³)") +
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, NA))

p2 <- ggplot(Data[is.na(Data$Density)==FALSE, ], aes(x = RW)) +
  geom_histogram(binwidth = 0.5, boundary = 0 , fill = "lightblue", color = "black") +
  facet_wrap(~species_temp) +
  labs(x = "Ring Width (mm)") +
  theme_bw() +
  theme(legend.position = "bottom") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, NA))

ggarrange(p1, p2, ncol = 1, heights = c(2, 1))

```

# average years per tree

```{r}

# average per year
Data <- Data %>%
  group_by(Year, TreeID, species_temp) %>%
  summarise(
    RW = if (all(is.na(RW))) NA_real_ else mean(RW, na.rm = TRUE),
    Density = if (all(is.na(Density))) NA_real_ else mean(Density, na.rm = TRUE),
    .groups = "drop"
  )

```

# fill missing density values

One GAM per species, fitted on all sites at once -- as in `CodeMain.qmd`, where
the imputation also ran before the Ghent filter. The density-ring-width
relationship is a wood property, not a site property, so pooling is deliberate:
it also means the imputation for a 7-tree site borrows the shape of the
Ghent curve instead of fitting one to almost nothing.

```{r}

# Fit GAM for each species
models <- Data %>%
  filter(!is.na(RW), !is.na(Density)) %>%
  group_by(species_temp) %>%
  group_split() %>%
  setNames(unique(Data$species_temp)) %>%
  lapply(function(dat) {
    gam(
      Density ~ s(RW, k = 6),
      data = dat,
      method = "REML"
    )
  })

# predict density for the missing values
Data <- Data %>%
  mutate(Density_imputed = is.na(Density)) %>%
  group_by(species_temp) %>%
  group_modify(~ {

    model <- models[[unique(.y$species_temp)]]

    .x$Density_predicted <- predict(
      model,
      newdata = .x
    )

    .x$Density <- ifelse(
      is.na(.x$Density) & !is.na(.x$RW),
      .x$Density_predicted,
      .x$Density
    )

    .x
  }) %>%
  ungroup()

# plot the models
ggplot(Data, aes(RW, Density)) +
  geom_point(
    aes(shape = Density_imputed),
    alpha = 0.5,
    size = 1
  ) +
  geom_smooth(
    data = Data %>% filter(!Density_imputed),
    method = "gam",
    formula = y ~ s(x, k = 6),
    se = TRUE
  ) +
  facet_wrap(~species_temp, scales = "free_y") +
  theme_bw()

# how much was imputed, per site
Data %>%
  mutate(Site = substr(TreeID, 1, 3)) %>%
  group_by(Site) %>%
  summarise(n = n(), imputed = sum(Density_imputed),
            pct_imputed = round(100 * mean(Density_imputed), 1),
            .groups = "drop")

# remove density_predicted
Data$Density_predicted <- NULL

```

# add metadata

```{r}

# load tree metadata, calculate diameter and barkless diameter and add to Data
metaData <- read_excel(metaDataPath) %>%
  rename(TreeID = `Tree`)

metaData <- metaData %>%
  mutate(DBH_mm = Circumference_Drilling_Height_mm / pi) %>%
  mutate(Barkless_DBH_mm = DBH_mm - Single_Bark_Thickness_mm * 2) %>%
  select(-c(Circumference_Drilling_Height_mm))

Data <- Data %>%
  left_join(metaData, by = c("TreeID"))

# remove trees that were not cored or did not crossdate
Data <- Data %>%
  filter(Cored == "YES", Crossdate_Status == "Yes") %>%
  select(-c(Cored, Crossdate_Status))

# remove non useful columns
Data <- Data %>%
  select(-c(Circumference_Breast_Height_AstridVDB_mm, Remarks, Last_Complete_Ring, Logger_Data_Present_AstridVandenbossche, Soil_Data_present_AstridVandenbossche, LIDAR_data_Present_KarlienMoeys, Dendrometer_Data_Present_KarenDePauw, Owner, Email, Phone, species_temp))


#### site column, and the check that makes it trustworthy ####

# `Site` is derived from the TreeID rather than from `Region`, so it cannot
# disagree with the ERA5 file it selects. The cross-check below is what
# catches a mislabelled tree: a TreeID whose prefix does not match its
# Region, or a site present in the data but absent from SITES.
Data <- Data %>%
  mutate(Site = substr(TreeID, 1, 3))

site_check <- Data %>%
  distinct(Site, Region) %>%
  full_join(SITES %>% select(Site, Region_cfg = Region), by = "Site")

if (any(is.na(site_check$Region) | is.na(site_check$Region_cfg) |
        site_check$Region != site_check$Region_cfg)) {
  print(site_check)
  stop("Site/Region mismatch: every site in the data needs exactly one row in ",
       "SITES with the same Region. Add the site (with its latitude and ",
       "elevation) or fix the TreeID prefix.")
}

# unify Site to a factor in SITES order, so every table and facet strip below
# reads west-to-east / south-to-north in the same order
Data <- Data %>%
  mutate(Site = factor(Site, levels = SITES$Site))

Data %>%
  group_by(Site, Region) %>%
  summarise(n_trees = n_distinct(TreeID), n_rings = n(),
            first_year = min(Year), last_year = max(Year), .groups = "drop")

```

# correct RW and density for shrinkage: species-specific total radial shrinkage for ring width, and species-specific total volumetric shrinkage for density. based on CIRAD Tropix sheets and GWDD database

```{r}

# Trees_Cooltree.xlsx spells two of the three species inconsistently
# ("Fraxinus exelsior" alongside "Fraxinus excelsior", "Tillia x europaea"
# alongside "Tilia x europaea"). With Ghent alone that never mattered; across
# six sites a typo would join to Sr = NA, propagate into RW_green, and surface
# only as bakker_bai()'s "Required columns contain NA values". Normalise
# first, then assert the join was complete.
Data <- Data %>%
  mutate(Species = recode(
    Species,
    "Fraxinus exelsior" = "Fraxinus excelsior",
    "Tillia x europaea" = "Tilia x europaea"
  ))

# Species-specific shrinkage coefficients, based on CIRAD Tropix sheets and GWDD
shrinkage <- tibble(
  Species = c(
    "Fraxinus excelsior",
    "Quercus robur",
    "Tilia x europaea"
  ),

  # total green -> oven-dry radial shrinkage
  Sr = c(
    0.057,
    0.045,
    0.050
  ),

  # volumetric shrinkage used for density correction
  Sv = c(
    0.1475,
    0.1376,
    0.1200
  )
)


Data <- Data %>%
  left_join(shrinkage, by = "Species") %>%
  mutate(

    # oven-dry ring width -> reconstructed green ring width
    RW_green = RW / (1 - Sr),

    # oven-dry density -> basic density
    # i.e. oven-dry mass / green volume
    Density_basic = Density * (1 - Sv)

  )

if (anyNA(Data$Sr)) {
  print(Data %>% filter(is.na(Sr)) %>% distinct(Site, Species))
  stop("Species without shrinkage coefficients: add them to `shrinkage` or ",
       "add the spelling to the recode above.")
}

Data <- Data %>% select(-c(Sr, Sv))

```

# Basal area geometry (Bakker method)

`BAI_mm2` survives only because the basal biomass increment of the next chunk is `BAI x basic density`; nothing downstream analyses it. From the chronologies onwards the response variable is BBI.

```{r}


bakker_bai <- function(data) {
  
  # Required columns
  required_cols <- c(
    "TreeID",
    "Year",
    "RW_green",
    "Barkless_DBH_mm",
    "Distance_to_Pith_mm"
  )
  
  missing_cols <- setdiff(required_cols, names(data))
  
  if (length(missing_cols) > 0) {
    stop(
      "Missing required columns: ",
      paste(missing_cols, collapse = ", ")
    )
  }
  
  # Check for missing values in variables needed for calculation
  if (anyNA(data[, required_cols])) {
    stop("Required columns contain NA values.")
  }
  
  # Check ring widths
  if (any(data$RW_green < 0)) {
    stop("RW_green cannot contain negative values.")
  }
  
  # Sort chronologically within tree
  data <- data[
    order(data$TreeID, data$Year),
  ]
  
  # Check that DBH and distance-to-pith are constant within tree
  dbh_n <- tapply(
    data$Barkless_DBH_mm,
    data$TreeID,
    function(x) length(unique(x))
  )
  
  d2p_n <- tapply(
    data$Distance_to_Pith_mm,
    data$TreeID,
    function(x) length(unique(x))
  )
  
  if (any(dbh_n > 1)) {
    stop("Barkless_DBH_mm must be constant within each TreeID.")
  }
  
  if (any(d2p_n > 1)) {
    stop("Distance_to_Pith_mm must be constant within each TreeID.")
  }
  
  # Split by tree
  out <- lapply(
    split(data, data$TreeID),
    function(x) {
      
      # Current inside-bark DBH
      dbh_current <- x$Barkless_DBH_mm[1]
      
      # Distance from innermost measured ring to estimated pith
      d2p <- x$Distance_to_Pith_mm[1]
      
      # Chronological radius represented by the core:
      #
      # estimated unmeasured distance to pith
      # +
      # all measured ring widths
      I_P <- d2p + sum(x$RW_green)
      
      if (I_P <= 0) {
        stop(
          "Total chronological radius must be > 0 for TreeID ",
          x$TreeID[1]
        )
      }
      
      # Radius accumulated BEFORE each measured ring
      radial_start <- d2p + c(
        0,
        head(cumsum(x$RW_green), -1)
      )
      
      # Radius accumulated through the END of each measured ring
      radial_end <- d2p + cumsum(x$RW_green)
      
      # Bakker proportional factors
      G_start <- radial_start / I_P
      G_end   <- radial_end   / I_P
      
      # Reconstructed inside-bark diameter
      DBH_start <- dbh_current * G_start
      DBH_end   <- dbh_current * G_end
      
      # Basal area (mm^2)
      BA_start <- pi * (DBH_start / 2)^2
      BA_end   <- pi * (DBH_end / 2)^2
      
      # Annual basal area increment
      BAI <- BA_end - BA_start
      
      # Add results
      x$BA_startyear_mm2 <- BA_start
      x$BAI_mm2 <- BAI
      
      x
    }
  )
  
  # Return long-format dataframe
  out <- do.call(rbind, out)
  rownames(out) <- NULL
  
  out
}


Data <- bakker_bai(Data)



# No BAI diagnostics here on purpose: BAI is an intermediate now, so the
# increment-against-size plots live in the BBI chunk below and nowhere else.


```

# BBI calculations

```{r}

# density is in kg/m³, we need to convert it to g/mm³
Data$Density_gmm3 <- Data$Density_basic / 1000000

# calculate BBI (basal biomass increment) in g/mm
Data$BBI_gmm <- Data$BAI_mm2 * Data$Density_gmm3




# plot
ggplot(Data, aes(x = BA_startyear_mm2, y = BBI_gmm)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(Species ~ Region) +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(xlim = c(0, 1000000), ylim = c(0, 30))

# per site, on its own scale: the non-Ghent sites are 5-8 trees each and
# would be invisible inside a shared axis range
ggplot(Data, aes(x = BA_startyear_mm2, y = BBI_gmm)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_wrap(~Site, scales = "free") +
  theme_bw() +
  theme(legend.position = "bottom")


```

determine cambial age and cut first years

```{r}

#| label: cambial-age

#' Add cambial age to a long-format tree-ring dataframe
#'
#' CambialAge_measured : ring index counting from the innermost MEASURED ring
#'                       (= 1 for that ring). This is exact but underestimates
#'                       true cambial age whenever the core missed the pith.
#'
#' PithOffset_est      : estimated number of unmeasured rings between pith and
#'                       innermost measured ring, from Distance_to_Pith_mm
#'                       divided by the mean width of the innermost `n_inner`
#'                       measured rings (sensu the linear-extrapolation
#'                       approach; cf. Duncan 1989 for the geometric
#'                       alternative based on arc curvature).
#'
#' CambialAge         : CambialAge_measured + PithOffset_est, rounded.

n_inner <- 10   # rings used to estimate the near-pith growth rate

Data <- Data %>%
  arrange(TreeID, Year) %>%
  group_by(TreeID) %>%
  mutate(
    CambialAge_measured = row_number(),
    
    # mean width of the innermost measured rings (green/shrinkage-corrected,
    # to match the units of Distance_to_Pith_mm)
    RW_inner_mean = mean(RW_green[CambialAge_measured <= n_inner], na.rm = TRUE),
    
    PithOffset_est = if_else(
      is.na(Distance_to_Pith_mm) | RW_inner_mean <= 0,
      NA_real_,
      round(Distance_to_Pith_mm / RW_inner_mean)
    ),
    
    CambialAge = CambialAge_measured + coalesce(PithOffset_est, 0),
    
  ) %>%
  ungroup() %>%
  select(-RW_inner_mean)


# how much of the collection is affected, and by how much
Data %>%
  group_by(Site, TreeID, Species) %>%
  summarise(
    d2p_mm    = first(Distance_to_Pith_mm),
    offset    = first(PithOffset_est),
    n_rings   = n(),
    first_yr  = min(Year),
    .groups   = "drop"
  ) %>%
  arrange(desc(offset)) %>%
  print(n = 20)


# cut first x years of each tree
nyears_cut <- 15
Data <- Data %>% 
  filter(CambialAge > nyears_cut)

# remove samples with less then x years left
nyears_min <- 5
Data <- Data %>%
  group_by(TreeID) %>%
  filter(n() >= nyears_min) %>%
  ungroup()


```

# detrend BBI

Primary: Cook & Peters (1997), The Holocene 7:361–370. This is the citation for transform-then-subtract. Log is their p = 0 case, so cite it as such rather than implying they recommend log unconditionally — something like “following Cook and Peters (1997), series were log-transformed and indices computed as residuals rather than ratios, avoiding the instability of the ratio index where the fitted curve approaches zero.”

```{r}

#' Detrend a long-format tree-ring dataframe with dplR
#'
#' @param data       A long-format data.frame/tibble with at least a year
#'   column, a sample/series id column, and the column to be detrended.
#' @param value_col  Name (string) of the column to detrend, e.g. "RW" or
#'   "Density".
#' @param year_col   Name (string) of the year column. Default "Year".
#' @param sample_col Name (string) of the sample/series id column.
#'   Default "Sample".
#' @param fill_method Method passed to `dplR::fill.internal.NA` for filling
#'   internal gaps before detrending (e.g. "Spline", "mean", "linear").
#'   Default "Spline".
#' @param detrend_method Method passed to `dplR::detrend`, e.g. "Spline",
#'   "ModNegExp", "Mean", "Ar", "Friedman", "AgeDepSpline". Default "Spline".
#' @param out_col    Name (string) of the new detrended column that is added
#'   to `data`. Default `paste0(value_col, "_Detrended")`.
#' @param ...        Extra arguments passed on to `dplR::detrend`
#'   (e.g. `nyrs = 50`, `f = 0.5`).
#'
#' @return `data` with one extra column (`out_col`) holding the detrended
#'   values. Rows where `value_col` was originally NA keep NA in `out_col`.
#'   Rows that fell inside a filled-in gap are dropped again after
#'   detrending, exactly like the manual pipeline in code_Main.Rmd.
#'
dplr_detrend_long <- function(data,
                              value_col,
                              year_col = "Year",
                              sample_col = "Sample",
                              fill_method = "Spline",
                              detrend_method = "Spline",
                              out_col = paste0(value_col, "_Detrended"),
                              ...) {
  
  stopifnot(value_col %in% names(data),
            year_col %in% names(data),
            sample_col %in% names(data))
  
  wide_input <- data[, c(year_col, sample_col, value_col)]
  names(wide_input) <- c("Year", "Sample", "Value")
  wide_input <- wide_input %>% filter(!is.na(Value))
  
  wide <- wide_input %>%
    pivot_wider(names_from = "Sample", values_from = "Value") %>%
    arrange(Year) %>%
    complete(Year = seq(min(Year), max(Year), 1))
  
  years <- wide$Year
  wide$Year <- NULL
  wide <- as.data.frame(wide, check.names = FALSE)   # <-- the fix
  row.names(wide) <- years
  
  wide_filled <- fill.internal.NA(wide, fill = fill_method)
  detrended <- detrend(wide_filled, method = detrend_method, ...)
  
  # print(rwi.stats(detrended))
  # plot(time(detrended), sss(detrended),type="l",ylim=c(0.4,1),
  #      col="darkblue",lwd=2,xlab="Year",ylab="SSS")
  # abline(h=0.85,col="blue",lty="dashed")
  
  detrended$Year <- as.numeric(row.names(detrended))
  detrended_long <- pivot_longer(
    detrended,
    cols = -"Year",
    names_to = "Sample",
    values_to = out_col
  )
  
  result <- full_join(
    data, detrended_long,
    by = setNames(c("Year", "Sample"), c(year_col, sample_col))
  )
  
  result <- result[!is.na(result[[value_col]]), ]
  
  result
}



# log transform 
Data$logBBI <- log(Data$BBI_gmm *1000) # BBI in g/m


# detrend BBI residuals
Data <- dplr_detrend_long(
  data = Data,
  year_col = "Year",
  sample_col = "TreeID", 
  value_col = "logBBI",
  detrend_method = "Spline",
  nyrs = 30,
  out_col = "logBBI_detrended", 
  difference = TRUE
)




```

# make chronologies for each site and species

`CodeMain.qmd` grouped by `Species` alone, after filtering to Ghent. Here the
grouping is `Site x Species`, and the replication floor is `MIN_REPL` (see the
configuration chunk) rather than a hard-coded 10.

```{r chronologies}

ChronsAll <- Data %>%
  group_by(Site, Region, Species, Year) %>%
  summarise(
    BBI_chron = tbrm(logBBI_detrended),
    BBI_upper = tbrm(logBBI_detrended) + (qnorm(0.975)*sd(logBBI_detrended)/sqrt(n())),
    BBI_lower = tbrm(logBBI_detrended) - (qnorm(0.975)*sd(logBBI_detrended)/sqrt(n())),
    BBI_n = sum(!is.na(logBBI_detrended)),
    .groups = "drop"
  )

# what replication is actually available, before the floor is applied. This is
# the table to look at when deciding MIN_REPL: `max_n` is the best any year at
# that site x species can do.
ChronsAll %>%
  group_by(Site, Species) %>%
  summarise(
    max_n       = max(BBI_n),
    yrs_ge_repl = sum(BBI_n >= MIN_REPL & Year >= FIRST_YEAR),
    # guarded, because a site x species that never reaches MIN_REPL would
    # otherwise give min(integer(0)) = Inf and a warning
    first_yr    = if (any(BBI_n >= MIN_REPL)) min(Year[BBI_n >= MIN_REPL]) else NA_integer_,
    last_yr     = if (any(BBI_n >= MIN_REPL)) max(Year[BBI_n >= MIN_REPL]) else NA_integer_,
    .groups = "drop"
  ) %>%
  print(n = 30)

# cut at the replication floor and at the start of the instrumental record
ChronsAll <- ChronsAll %>%
  filter(BBI_n >= MIN_REPL, Year >= FIRST_YEAR)

# Drop site x species chronologies too short to reach MIN_N overlapping years
# in any window. Leaving them in produces an all-NA surface and then a
# "No rows for Chron = ..." error inside the plotting function, several
# minutes into the loop.
short <- ChronsAll %>%
  count(Site, Species) %>%
  filter(n < MIN_YEARS)

if (nrow(short)) {
  message("Dropping ", nrow(short), " chronolog", if (nrow(short) == 1) "y" else "ies",
          " shorter than MIN_YEARS = ", MIN_YEARS, ":")
  print(short)
  ChronsAll <- ChronsAll %>% anti_join(short, by = c("Site", "Species"))
}

ChronsAll %>%
  group_by(Site, Species) %>%
  summarise(n_years = n(), first_yr = min(Year), last_yr = max(Year),
            mean_n = round(mean(BBI_n), 1), .groups = "drop") %>%
  print(n = 30)


#### figure: every chronology, one row per site ####

chronplot <- ggplot(ChronsAll, aes(x = Year)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_line(aes(y = BBI_chron), colour = "red", linewidth = 0.4) +
  geom_ribbon(aes(ymin = BBI_lower, ymax = BBI_upper), fill = "red", alpha = 0.2) +
  facet_grid(Site ~ Species) +
  labs(x = "Year", y = "Detrended BBI chronology") +
  theme_bw(base_size = 9) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey92", colour = NA)) +
  scale_x_continuous(expand = c(0, 0))

replication_plot <- ggplot(ChronsAll, aes(x = Year)) +
  geom_bar(aes(y = BBI_n), stat = "identity", fill = "red", alpha = 0.5) +
  facet_grid(Site ~ Species, scales = "free_y") +
  labs(x = "Year", y = "Replication (trees)") +
  theme_bw(base_size = 9) +
  theme(strip.background = element_rect(fill = "grey92", colour = NA)) +
  scale_x_continuous(expand = c(0, 0))

chronplot
replication_plot

ggsave(file.path(outRoot, "0_chronologies_allsites.png"), chronplot,
       width = 11, height = 12, dpi = 600)
ggsave(file.path(outRoot, "0_replication_allsites.png"), replication_plot,
       width = 11, height = 12, dpi = 600)

# the chronologies themselves, for reuse
write_csv(ChronsAll, file.path(outRoot, "Chrons_AllSites.csv"))
saveRDS(ChronsAll, file.path(outRoot, "Chrons_AllSites.rds"))

```

# Moving-window climate correlations

*The section heading, the prose and every function below are copied
verbatim from `CodeMain.qmd`. The site loop that drives them is in the
sections after this one.*

Every combination of **window length** (20-100 days) and **window end day** (1 May of the previous year to 30 November of the growth year) is aggregated from the daily ERA5 series -- totals for precipitation and for the water balance, means for Tmean, Tmax and Tmin -- and the resulting one-value-per-year series is correlated with each species chronology. Each window is used twice:

- **raw** -- the window as aggregated;
- **detrended** -- the same yearly series after a 30-year spline (`dplR::ffcsaps`, `f = 0.5`, i.e. the spline the tree-ring series were detrended with) has been *subtracted*. Subtraction rather than division, because a ratio index is meaningless for a temperature in degrees C (the spline crosses zero) and unstable for a 20-day precipitation total (which can be 0 mm).

A fifth variable, the **daily water balance** `D = P - PET`, is derived from the same daily series (section 1b below). Accumulated over a window it is the quantity SPEI is built on, so it doubles as the drought index used for extreme-year selection at the end of this document. Precipitation alone misses the demand side: a wet but hot August is not the same growing condition as a wet and cool one.

# functions

```{r}



#### ==================================================================== ####
####  1. Hourly ERA5 -> daily climate                                     ####
#### ==================================================================== ####

#' Aggregate hourly ERA5 to the daily fields the pipeline needs
#'
#' Two kinds of ERA5 field are handled differently:
#'
#'   INSTANTANEOUS (`t2m`, `d2m`, `u10`, `v10`, `sp`) belong to the calendar
#'     day they are stamped with.
#'   ACCUMULATED (`tp`, `ssrd`) are totals over the PRECEDING hour, so the
#'     value stamped 00:00 belongs to the previous calendar day and is
#'     shifted back one hour before the daily total.
#'
#' `t2m` and `tp` are required. The four Penman-Monteith fields are optional:
#' whichever are present get aggregated, and `add_water_balance()` reports
#' clearly if any are missing.
#'
#' @param clim Data frame with `valid_time` (parseable date-time) plus:
#'   REQUIRED  `t2m` 2 m air temperature, K
#'             `tp`  total precipitation, m per hour
#'   OPTIONAL  `d2m`  2 m dewpoint temperature, K
#'             `u10`  10 m eastward wind, m/s
#'             `v10`  10 m northward wind, m/s
#'             `ssrd` surface solar radiation downwards, J/m2 per hour
#'             `sp`   surface pressure, Pa
#' @param tp_hour_ending TRUE (default) applies the one-hour shift described
#'   above to every accumulated field; FALSE aggregates 00:00-23:00 as stamped.
#' @param require_full_day Only keep days with all 24 hours present. A day
#'   with a gap becomes NA, which in turn invalidates every window that
#'   contains it (see `window_matrix`) rather than silently shortening it.
#' @param wind_height Height of the ERA5 wind field in m (10 for `u10`/`v10`).
#'
#' @return Tibble, one row per calendar day over a gap-free date sequence:
#'   `Date`, `Tmean`, `Tmax`, `Tmin` (degrees C), `P` (mm/day), and where the
#'   inputs allowed: `Tdew` (degrees C), `u2` (m/s at 2 m), `Rs` (MJ/m2/day),
#'   `Pkpa` (kPa).
era5_to_daily <- function(clim,
                          tp_hour_ending  = TRUE,
                          require_full_day = TRUE,
                          wind_height     = 10) {

  stopifnot(all(c("valid_time", "t2m", "tp") %in% names(clim)))

  dt <- clim$valid_time
  if (!inherits(dt, "POSIXct")) {
    dt <- as.POSIXct(as.character(dt), tz = "UTC")
  }
  if (anyNA(dt)) stop("valid_time could not be parsed for all rows.")

  has <- function(v) v %in% names(clim)

  hourly <- tibble::tibble(
    DateI = as.Date(dt, tz = "UTC"),                                  # instant
    DateA = as.Date(dt - if (tp_hour_ending) 3600 else 0, tz = "UTC") # accum
  )

  # ---- instantaneous: air temperature (required) ----------------------
  hourly$t2m <- as.numeric(clim$t2m)
  parts <- list(
    hourly %>%
      dplyr::group_by(Date = .data$DateI) %>%
      dplyr::summarise(
        n_t   = sum(!is.na(.data$t2m)),
        Tmean = mean(.data$t2m, na.rm = TRUE) - 273.15,
        Tmax  = suppressWarnings(max(.data$t2m, na.rm = TRUE)) - 273.15,
        Tmin  = suppressWarnings(min(.data$t2m, na.rm = TRUE)) - 273.15,
        .groups = "drop"
      )
  )

  # ---- accumulated: precipitation (required) --------------------------
  hourly$tp <- as.numeric(clim$tp)
  parts <- c(parts, list(
    hourly %>%
      dplyr::group_by(Date = .data$DateA) %>%
      dplyr::summarise(
        n_p = sum(!is.na(.data$tp)),
        P   = sum(.data$tp, na.rm = TRUE) * 1000,      # m -> mm
        .groups = "drop"
      )
  ))

  # ---- instantaneous: dewpoint ---------------------------------------
  if (has("d2m")) {
    hourly$d2m <- as.numeric(clim$d2m)
    parts <- c(parts, list(
      hourly %>%
        dplyr::group_by(Date = .data$DateI) %>%
        dplyr::summarise(
          n_d  = sum(!is.na(.data$d2m)),
          Tdew = mean(.data$d2m, na.rm = TRUE) - 273.15,
          .groups = "drop"
        )
    ))
  }

  # ---- instantaneous: wind -------------------------------------------
  # The MAGNITUDE is computed hourly and only then averaged. Averaging the
  # components first and taking the magnitude of the mean vector cancels out
  # opposing directions and underestimates wind speed, sometimes badly.
  # FAO-56 eq. 47 converts the measurement height to the 2 m reference.
  if (has("u10") && has("v10")) {
    hourly$wspd <- sqrt(as.numeric(clim$u10)^2 + as.numeric(clim$v10)^2)
    k2 <- 4.87 / log(67.8 * wind_height - 5.42)
    parts <- c(parts, list(
      hourly %>%
        dplyr::group_by(Date = .data$DateI) %>%
        dplyr::summarise(
          n_w = sum(!is.na(.data$wspd)),
          u2  = mean(.data$wspd, na.rm = TRUE) * k2,
          .groups = "drop"
        )
    ))
  }

  # ---- accumulated: incoming shortwave radiation ----------------------
  if (has("ssrd")) {
    hourly$ssrd <- as.numeric(clim$ssrd)
    parts <- c(parts, list(
      hourly %>%
        dplyr::group_by(Date = .data$DateA) %>%
        dplyr::summarise(
          n_r = sum(!is.na(.data$ssrd)),
          Rs  = sum(.data$ssrd, na.rm = TRUE) / 1e6,   # J/m2 -> MJ/m2/day
          .groups = "drop"
        )
    ))
  }

  # ---- instantaneous: surface pressure --------------------------------
  if (has("sp")) {
    hourly$sp <- as.numeric(clim$sp)
    parts <- c(parts, list(
      hourly %>%
        dplyr::group_by(Date = .data$DateI) %>%
        dplyr::summarise(
          n_s  = sum(!is.na(.data$sp)),
          Pkpa = mean(.data$sp, na.rm = TRUE) / 1000,  # Pa -> kPa
          .groups = "drop"
        )
    ))
  }

  # ---- stitch onto a gap-free calendar --------------------------------
  all_dates <- sort(unique(unlist(lapply(parts, function(p) p$Date))))
  daily <- tibble::tibble(
    Date = seq(as.Date(min(all_dates), origin = "1970-01-01"),
               as.Date(max(all_dates), origin = "1970-01-01"), by = "day")
  )
  for (p in parts) daily <- dplyr::left_join(daily, p, by = "Date")

  count_cols <- intersect(c("n_t", "n_p", "n_d", "n_w", "n_r", "n_s"),
                          names(daily))
  daily <- daily %>%
    dplyr::mutate(dplyr::across(dplyr::all_of(count_cols),
                                ~ dplyr::coalesce(.x, 0L)))

  # A value is only kept if its OWN field had all 24 hours that day; each
  # field is masked by its own counter, so one missing radiation hour does
  # not throw away that day's temperature.
  if (require_full_day) {
    mask <- function(d, cols, cnt) {
      cols <- intersect(cols, names(d))
      if (!length(cols) || !cnt %in% names(d)) return(d)
      keep <- d[[cnt]] == 24L
      d[cols] <- lapply(d[cols],
                        function(x) dplyr::if_else(keep, x, NA_real_))
      d
    }
    daily <- daily %>%
      mask(c("Tmean", "Tmax", "Tmin"), "n_t") %>%
      mask("P",    "n_p") %>%
      mask("Tdew", "n_d") %>%
      mask("u2",   "n_w") %>%
      mask("Rs",   "n_r") %>%
      mask("Pkpa", "n_s")
  }

  incomplete <- Reduce(`|`, lapply(count_cols, function(cc) daily[[cc]] != 24L))
  if (sum(incomplete) > 0) {
    message("era5_to_daily(): ", sum(incomplete),
            " day(s) incomplete in at least one field (of ", nrow(daily), ").")
  }
  present <- setdiff(names(daily), c("Date", count_cols))
  message("era5_to_daily(): daily fields -> ", paste(present, collapse = ", "))

  daily %>% dplyr::select(-dplyr::all_of(count_cols))
}


#### ==================================================================== ####
####  2. Reference evapotranspiration and the daily water balance        ####
#### ==================================================================== ####
####                                                                      ####
####  FAO-56 Penman-Monteith (Allen et al. 1998), the standard method     ####
####  and the one SPEIbase v2.0 uses. Implemented directly rather than    ####
####  via SPEI::penman(), because that function is written for MONTHLY    ####
####  series: it treats a plain vector as consecutive months starting in  ####
####  January and computes Ra per month. Handing it daily data does not   ####
####  error -- it silently cycles the days through the twelve months and  ####
####  returns nonsense.                                                   ####
####                                                                      ####
####  Validated against FAO-56 Chapter 4, Example 18 (Uccle, Belgium,     ####
####  50.80 N, 100 m, 6 July) -- see fao56_example18().              ####
#### ==================================================================== ####


#' Saturation vapour pressure at a given temperature
#'
#' FAO-56 equation 11. @param Ta Temperature, degrees C. @return kPa.
#' @noRd
.svp <- function(Ta) 0.6108 * exp(17.27 * Ta / (Ta + 237.3))


#' Extraterrestrial radiation
#'
#' FAO-56 equations 21-25. Ra depends only on latitude and day of year, so it
#' carries the seasonal cycle of available energy at the top of the atmosphere
#' with no measurements at all. Needed here for the clear-sky reference Rso.
#'
#' @param lat Latitude in decimal degrees (negative south).
#' @param doy Day of year (1-366).
#' @return Ra in MJ m-2 day-1.
#' @noRd
.extraterrestrial_radiation <- function(lat, doy) {

  phi <- lat * pi / 180

  # inverse relative Earth-Sun distance and solar declination
  dr    <- 1 + 0.033 * cos(2 * pi * doy / 365)
  delta <- 0.409 * sin(2 * pi * doy / 365 - 1.39)

  # sunset hour angle; clamped for polar day / polar night
  ws <- acos(pmin(pmax(-tan(phi) * tan(delta), -1), 1))

  # Gsc = 0.0820 MJ m-2 min-1
  (24 * 60 / pi) * 0.0820 * dr *
    (ws * sin(phi) * sin(delta) + cos(phi) * cos(delta) * sin(ws))
}


#' FAO-56 Penman-Monteith reference evapotranspiration, daily time step
#'
#'   ET0 = [ 0.408 D (Rn - G) + g 900/(T+273) u2 (es - ea) ]
#'         / [ D + g (1 + 0.34 u2) ]                        (FAO-56 eq. 6)
#'
#' G is taken as zero, which FAO-56 recommends for a 24-hour step.
#'
#' @param Tmean,Tmax,Tmin Daily temperatures, degrees C.
#' @param ea   Actual vapour pressure, kPa.
#' @param u2   Wind speed at 2 m, m/s.
#' @param Rs   Incoming shortwave radiation, MJ m-2 day-1.
#' @param Pkpa Atmospheric pressure, kPa.
#' @param doy  Day of year.
#' @param lat  Latitude, decimal degrees.
#' @param elev Elevation, m a.s.l.
#' @return ET0 in mm/day, floored at zero.
penman_monteith_pet <- function(Tmean, Tmax, Tmin, ea, u2, Rs, Pkpa,
                                doy, lat, elev) {

  stopifnot(length(lat) == 1, length(elev) == 1,
            is.finite(lat), is.finite(elev))

  # --- vapour pressure -------------------------------------------------
  es <- (.svp(Tmax) + .svp(Tmin)) / 2                      # eq. 12
  ea <- pmin(ea, es)          # ea > es means supersaturation; cap it
  vpd <- pmax(es - ea, 0)

  # --- radiation -------------------------------------------------------
  Ra  <- .extraterrestrial_radiation(lat, doy)
  Rso <- (0.75 + 2e-5 * elev) * Ra                         # eq. 37
  Rns <- (1 - 0.23) * Rs      # albedo 0.23 = FAO reference grass; eq. 38

  # eq. 39. The cloudiness factor 1.35 Rs/Rso - 0.35 is only calibrated for
  # Rs/Rso >= 0.3; below that it turns Rnl negative, which would inflate Rn.
  # Rso also collapses toward zero in midwinter, so the ratio is clamped.
  ratio <- pmin(pmax(Rs / Rso, 0.3), 1)
  Rnl <- 4.903e-9 *
    (((Tmax + 273.16)^4 + (Tmin + 273.16)^4) / 2) *
    (0.34 - 0.14 * sqrt(pmax(ea, 0))) *
    (1.35 * ratio - 0.35)

  Rn <- Rns - Rnl
  G  <- 0                                                  # eq. 42

  # --- psychrometrics --------------------------------------------------
  delta <- 4098 * .svp(Tmean) / (Tmean + 237.3)^2          # eq. 13
  gamma <- 0.000665 * Pkpa                                 # eq. 8

  et0 <- (0.408 * delta * (Rn - G) +
            gamma * 900 / (Tmean + 273) * u2 * vpd) /
    (delta + gamma * (1 + 0.34 * u2))

  pmax(et0, 0)
}


#' Self-test: reproduce FAO-56 Chapter 4, Example 18
#'
#' Uccle (Brussels), 50.80 N, 100 m a.s.l., 6 July. Published inputs are
#' Tmax 21.5, Tmin 12.3 degrees C, ea 1.409 kPa, u2 2.078 m/s (10 km/h at
#' 10 m), Rs 22.07 MJ m-2 d-1, P 100.11 kPa; published answer ETo = 3.9 mm/day.
#'
#' Same latitude as Ghent, so this is a meaningful check of the whole
#' radiation path and not just the arithmetic. Run it once after any edit.
#'
#' @return Invisibly, the computed ET0. Prints a pass/fail line.
fao56_example18 <- function() {

  et0 <- penman_monteith_pet(
    Tmean = (21.5 + 12.3) / 2, Tmax = 21.5, Tmin = 12.3,
    ea = 1.409, u2 = 2.078, Rs = 22.07, Pkpa = 100.11,
    doy = 187, lat = 50.80, elev = 100
  )

  ok <- abs(et0 - 3.9) < 0.05
  message(sprintf("FAO-56 Example 18: ET0 = %.3f mm/day (published 3.9) -- %s",
                  et0, if (ok) "PASS" else "FAIL"))
  if (!ok) warning("penman_monteith_pet() does not reproduce FAO-56 Ex. 18.")
  invisible(et0)
}


#' Append PET and the daily water balance to a daily climate table
#'
#' `D = P - PET` is signed, in mm/day. Accumulated over a window (agg = "sum")
#' it is the climatic water balance -- the quantity SPEI standardises.
#'
#' @param daily Output of `era5_to_daily()`. Must carry `Tdew`, `u2`, `Rs`
#'   and `Pkpa` as well as the temperatures and `P`.
#' @param lat  Latitude of the grid cell, decimal degrees. Ghent is 51.05.
#'   Use each region's own latitude -- Ra changes by roughly 10 % per 5
#'   degrees, so one shared value biases PET for the more distant regions.
#' @param elev Elevation, m a.s.l. Affects Rso and the psychrometric
#'   constant; both are weak, so a rough figure is fine.
#' @return `daily` with two extra columns, `PET` and `D`.
add_water_balance <- function(daily, lat, elev) {

  needed <- c("Date", "Tmean", "Tmax", "Tmin", "P", "Tdew", "u2", "Rs", "Pkpa")
  missing <- setdiff(needed, names(daily))
  if (length(missing)) {
    stop("add_water_balance(): missing daily field(s): ",
         paste(missing, collapse = ", "),
         ".\n  Penman-Monteith needs ERA5 2m_dewpoint_temperature (d2m),",
         " 10m_u/v_component_of_wind (u10/v10),",
         " surface_solar_radiation_downwards (ssrd) and surface_pressure (sp)",
         " in the CSV passed to era5_to_daily().")
  }

  # UNIT GUARD. The (Tmean + 17.8)-style terms in FAO-56 are not
  # scale-invariant, so Kelvin would inflate PET by roughly a factor of ten.
  # `era5_to_daily()` already converts; this catches the function being
  # pointed at a raw t2m column by mistake.
  if (isTRUE(mean(daily$Tmean, na.rm = TRUE) > 100)) {
    stop("add_water_balance(): Tmean looks like Kelvin (mean > 100). ",
         "FAO-56 needs degrees C -- pass the output of era5_to_daily().")
  }

  daily %>%
    dplyr::mutate(
      PET = penman_monteith_pet(
        Tmean = .data$Tmean, Tmax = .data$Tmax, Tmin = .data$Tmin,
        ea    = .svp(.data$Tdew),      # eq. 14, from mean daily dewpoint
        u2    = .data$u2,
        Rs    = .data$Rs,
        Pkpa  = .data$Pkpa,
        doy   = as.integer(format(.data$Date, "%j")),
        lat   = lat, elev = elev
      ),
      D = .data$P - .data$PET
    )
}


#### ==================================================================== ####
####  2b. Standardisation, for extreme-year selection only               ####
#### ==================================================================== ####

#' Standardise an accumulated series to an SPEI-style z-score
#'
#' Nonparametric standardisation: empirical cumulative probabilities from
#' Gringorten plotting positions, mapped through the standard normal quantile
#' function. Same idea as SPEI -- accumulate, then map to a normal -- but
#' without fitting a log-logistic distribution, which can fail or behave
#' badly on short records and on windows that are almost always negative.
#'
#' Read exactly like SPEI: mean 0, sd ~1, below -1.5 conventionally "severe",
#' below -2 "extreme".
#'
#' Only used for extreme-year SELECTION. The response surfaces do not need
#' it: the transform is monotone, so it barely moves a Pearson r, and doing
#' it properly would mean one distribution fit per window length x end day.
#'
#' @param x Numeric vector, one accumulated value per year (may contain NA).
#' @param min_obs Return all-NA if fewer finite values than this.
#' @return Numeric vector the same length as `x`.
standardise_index <- function(x, min_obs = 30) {

  out <- rep(NA_real_, length(x))
  ok  <- is.finite(x)
  n   <- sum(ok)
  if (n < min_obs) {
    warning("standardise_index(): only ", n, " finite values; returning NA.")
    return(out)
  }

  # Gringorten plotting position: (rank - 0.44) / (n + 0.12)
  p <- (rank(x[ok], ties.method = "average") - 0.44) / (n + 0.12)
  out[ok] <- stats::qnorm(p)
  out
}


#' Build a standardised drought index for ONE fixed window
#'
#' @param daily Output of `add_water_balance()`.
#' @param years Growth years.
#' @param var   Climate column to accumulate; "D" for water balance.
#' @param end_offset,win_len The window, in `window_matrix()` offsets.
#' @param detrend Remove a 30-year spline before standardising. FALSE scores
#'   years against a fixed baseline (did this year actually stress the
#'   trees?); TRUE scores them against the contemporary norm (was this year
#'   unusual for its time?). See the note in the extremes chunk.
#' @return Tibble: Year, Value (accumulated), Index (standardised).
drought_index <- function(daily, years, var = "D",
                          end_offset, win_len,
                          detrend = FALSE, nyrs = 30) {

  window_series(daily, var, agg = "sum", years = years,
                end_offset = end_offset, win_len = win_len,
                detrend = detrend, nyrs = nyrs,
                detrend_method = "difference") %>%
    dplyr::mutate(Index = standardise_index(.data$Value))
}


#### ==================================================================== ####
####  3. Window aggregation                                              ####
#### ==================================================================== ####

# Windows are defined by (length in days, end day). The end day is an OFFSET
# in days from 1 January of the growth year:
#
#     offset    0  = 1 January of the growth year
#     offset   -1  = 31 December of the previous year
#     offset  180  = ~30 June of the growth year
#
# A window of length L ending at offset d covers offsets d-L+1 ... d, so it can
# straddle the turn of the year. In leap years a given offset falls one
# calendar day earlier after 29 February; that 1-day jitter is negligible
# against the 20-day minimum window.

#' Cumulative sums that survive NAs
#'
#' Returns padded cumulative sums of the values and of the valid-value count,
#' so any window aggregate is two subtractions and its completeness is known.
#' @noRd
.clim_cumsum <- function(x) {
  ok <- !is.na(x)
  list(
    s = c(0, cumsum(ifelse(ok, x, 0))),
    n = c(0L, cumsum(as.integer(ok)))
  )
}

#' Aggregate one climate variable over one window length, for many end days
#'
#' @param daily  Output of `era5_to_daily()` (gap-free daily date sequence).
#' @param var    Column of `daily` to aggregate.
#' @param agg    "sum" (precipitation totals) or "mean" (temperatures - note
#'   that Tmax over a window is the MEAN of the daily maxima, the usual
#'   dendroclimatological convention, not the single hottest day).
#' @param years  Growth years to compute the window for.
#' @param end_offsets  Window end days, as offsets from 1 Jan (see above).
#' @param win_len      Window length in days.
#'
#' @return Numeric matrix, `years` x `end_offsets`. A window is NA unless all
#'   `win_len` days fall inside the climate record and are complete.
window_matrix <- function(daily, var, agg = c("mean", "sum"),
                          years, end_offsets, win_len) {

  agg <- match.arg(agg)
  stopifnot(var %in% names(daily), win_len >= 1)

  cum <- .clim_cumsum(daily[[var]])
  n_days <- nrow(daily)
  day0 <- daily$Date[1]

  # Index into `daily` of 1 January of each growth year.
  jan1 <- as.integer(as.Date(paste0(years, "-01-01")) - day0) + 1L

  i_end   <- as.vector(outer(jan1, as.integer(end_offsets), "+"))
  i_start <- i_end - win_len + 1L

  inside <- i_start >= 1L & i_end <= n_days
  is_ <- pmin(pmax(i_start, 1L), n_days)
  ie_ <- pmin(pmax(i_end,   1L), n_days)

  s <- cum$s[ie_ + 1L] - cum$s[is_]
  n <- cum$n[ie_ + 1L] - cum$n[is_]

  out <- if (agg == "sum") s else s / n
  out[!inside | n < win_len] <- NA_real_

  matrix(out, nrow = length(years),
         dimnames = list(as.character(years), as.character(end_offsets)))
}

#' The yearly series of a single window (for plotting / inspection)
#'
#' @inheritParams window_matrix
#' @param end_offset Single window end day.
#' @param detrend Apply `spline_detrend()` to the yearly series.
#' @return Tibble with `Year` and `Value`.
window_series <- function(daily, var, agg, years, end_offset, win_len,
                          detrend = FALSE, nyrs = 30,
                          detrend_method = "difference") {

  v <- as.vector(window_matrix(daily, var, agg, years, end_offset, win_len))
  if (detrend) v <- spline_detrend(v, nyrs = nyrs, method = detrend_method)
  tibble::tibble(Year = years, Value = v)
}


#### ==================================================================== ####
####  4. Detrending the climate windows                                  ####
#### ==================================================================== ####

#' Remove a 30-year cubic smoothing spline from a yearly series
#'
#' Same spline as `dplR::detrend(method = "Spline", nyrs = 30)` applies to the
#' tree-ring series (`dplR::ffcsaps`, 50 % frequency response at 30 years), so
#' the climate series keeps the same frequency band as the chronologies it is
#' correlated with.
#'
#' The tree-ring pipeline divides by the spline; climate is detrended by
#' SUBTRACTION by default. A ratio index is meaningless for temperature in
#' degrees C (the curve crosses zero) and unstable for short-window
#' precipitation totals (a dry 20-day window can be 0 mm).
#'
#' @param y      Yearly series (may contain NAs at the ends).
#' @param nyrs   Spline stiffness in years.
#' @param method "difference" (y - spline) or "ratio" (y / spline).
#' @param min_obs Series shorter than this are returned as all-NA.
#' @return Numeric vector the same length as `y`.
spline_detrend <- function(y, nyrs = 30, method = c("difference", "ratio"),
                           min_obs = 15) {

  method <- match.arg(method)
  out <- rep(NA_real_, length(y))
  ok <- is.finite(y)
  if (sum(ok) < min_obs) return(out)

  curve <- dplR::ffcsaps(y = y[ok], x = which(ok), nyrs = nyrs, f = 0.5)

  out[ok] <- if (method == "ratio") {
    ifelse(abs(curve) < .Machine$double.eps^0.5, NA_real_, y[ok] / curve)
  } else {
    y[ok] - curve
  }
  out
}


#### ==================================================================== ####
####  5. The moving-window correlations                                   ####
#### ==================================================================== ####

#' Pearson r and pairwise-complete n between every column of X and of Y
#' @noRd
.cor_cols <- function(X, Y) {
  r <- suppressWarnings(stats::cor(X, Y, use = "pairwise.complete.obs"))
  n <- crossprod((!is.na(X)) * 1, (!is.na(Y)) * 1)
  list(r = r, n = n)
}

#' Two-sided p-value of a Pearson r
#' @noRd
.cor_p <- function(r, n) {
  df <- n - 2
  rr <- pmin(abs(r), 1 - 1e-12)
  t <- rr * sqrt(df / (1 - rr^2))
  p <- 2 * stats::pt(-abs(t), df = df)
  p[!is.finite(r) | df < 1] <- NA_real_
  p
}

#' Correlate chronologies against daily climate in moving windows
#'
#' Every combination of window length and window end day is aggregated from
#' the daily climate (means for temperature, totals for precipitation), giving
#' one value per growth year, and that yearly series is correlated with each
#' chronology. Each window is used twice: as aggregated ("raw") and after a
#' 30-year spline is removed ("detrended").
#'
#' @param chron    Long-format chronology table (one row per group x year).
#' @param daily    Output of `era5_to_daily()`.
#' @param chron_cols Chronology columns to correlate, e.g.
#'   `c(BBI = "BBI_chron")`. Names become labels.
#' @param group_col Column separating chronologies, e.g. "Species".
#' @param year_col  Year column.
#' @param clim_vars Named vector: names are `daily` columns, values the
#'   aggregation ("sum" or "mean").
#' @param win_lengths Window lengths in days.
#' @param end_offsets Window end days as offsets from 1 Jan of the growth year.
#' @param nyrs, detrend_method Passed to `spline_detrend()`.
#' @param min_n   Drop windows with fewer than this many overlapping years.
#' @param progress Print a line per climate variable.
#'
#' @return Tibble with one row per (group, chronology, variable, mode, window):
#'   `Group`, `Chron`, `Climate`, `Mode`, `WinLen`, `EndOffset`, `StartOffset`,
#'   `r`, `n`, `p`, `p_fdr` (Benjamini-Hochberg within each
#'   group x chronology x variable x mode surface).
moving_window_cor <- function(chron,
                              daily,
                              chron_cols  = c(BBI = "BBI_chron"),
                              group_col   = "Species",
                              year_col    = "Year",
                              clim_vars   = c(P = "sum", Tmean = "mean",
                                              Tmax = "mean", Tmin = "mean",
                                              D = "sum"),
                              win_lengths = seq(20, 100, by = 5),
                              end_offsets = seq(-245, 334),
                              nyrs = 30,
                              detrend_method = "difference",
                              min_n = 20,
                              progress = TRUE) {

  chron <- as.data.frame(chron)
  stopifnot(all(c(group_col, year_col, chron_cols) %in% names(chron)),
            all(names(clim_vars) %in% names(daily)))
  if (is.null(names(chron_cols))) names(chron_cols) <- chron_cols

  years <- sort(unique(chron[[year_col]]))
  groups <- sort(unique(as.character(chron[[group_col]])))

  # Response matrix: years x (group x chronology).
  series <- expand.grid(Group = groups, Chron = names(chron_cols),
                        stringsAsFactors = FALSE)
  Y <- matrix(NA_real_, nrow = length(years), ncol = nrow(series))
  for (j in seq_len(nrow(series))) {
    sel <- as.character(chron[[group_col]]) == series$Group[j]
    Y[match(chron[[year_col]][sel], years), j] <-
      chron[[chron_cols[[series$Chron[j]]]]][sel]
  }

  n_off <- length(end_offsets)
  out <- vector("list", length(clim_vars) * length(win_lengths) * 2L)
  k <- 0L

  for (v in names(clim_vars)) {
    if (progress) message("moving_window_cor(): ", v, " ...")
    for (L in win_lengths) {

      M <- window_matrix(daily, v, clim_vars[[v]], years, end_offsets, L)
      D <- apply(M, 2, spline_detrend, nyrs = nyrs, method = detrend_method)
      if (!is.matrix(D)) D <- matrix(D, nrow = nrow(M), ncol = ncol(M))

      for (md in c("raw", "detrended")) {
        st <- .cor_cols(if (md == "raw") M else D, Y)
        k <- k + 1L
        out[[k]] <- tibble::tibble(
          Group       = rep(series$Group, each = n_off),
          Chron       = rep(series$Chron, each = n_off),
          Climate     = v,
          Mode        = md,
          WinLen      = L,
          EndOffset   = rep(as.integer(end_offsets), times = nrow(series)),
          r           = as.vector(st$r),
          n           = as.vector(st$n)
        )
      }
    }
  }

  res <- dplyr::bind_rows(out) %>%
    dplyr::filter(!is.na(.data$r), .data$n >= min_n) %>%
    dplyr::mutate(
      StartOffset = .data$EndOffset - .data$WinLen + 1L,
      Climate = factor(.data$Climate, levels = names(clim_vars)),
      Mode    = factor(.data$Mode, levels = c("raw", "detrended")),
      p       = .cor_p(.data$r, .data$n)
    ) %>%
    dplyr::group_by(.data$Group, .data$Chron, .data$Climate, .data$Mode) %>%
    dplyr::mutate(p_fdr = stats::p.adjust(.data$p, method = "BH")) %>%
    dplyr::ungroup()

  attr(res, "n_years") <- range(res$n)
  res
}

#' Strongest window per group / chronology / variable / mode
#'
#' @param res Output of `moving_window_cor()`.
#' @param by  Grouping for the maximum.
#' @param n   Windows to keep per group.
best_windows <- function(res,
                         by = c("Group", "Chron", "Climate", "Mode"),
                         n = 1) {
  res %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(by))) %>%
    dplyr::slice_max(abs(.data$r), n = n, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(dplyr::across(dplyr::all_of(by)))
}


#### ==================================================================== ####
#### ==================================================================== ####
####  6. Figures                                                         ####
#### ==================================================================== ####

CLIM_LABELS <- c(
  P     = "Precipitation (sum)",
  Tmean = "Mean temperature",
  Tmax  = "Max temperature",
  Tmin  = "Min temperature",
  D     = "Water balance (P - PET, sum)"
)

CLIM_UNITS <- c(P = "mm", Tmean = "°C", Tmax = "°C", Tmin = "°C",
                D = "mm")

#' Month-start breaks for the window-end axis
#'
#' The axis is in days from 1 January of the growth year, so labels are taken
#' from a non-leap reference year; months of the previous year get a "*".
#'
#' @param every Label every n-th month (1 = all, 2 = Jan/Mar/May/..., 3 =
#'   Jan/Apr/Jul/Oct). Anchored on January so the turn of the year is always
#'   labelled.
#' @noRd
.offset_axis <- function(offsets, ref_year = 2003, every = 1) {
  rng <- range(offsets)
  sq <- seq(rng[1], rng[2])
  d <- as.Date(paste0(ref_year, "-01-01")) + sq
  keep <- as.integer(format(d, "%d")) == 1L &
    (as.integer(format(d, "%m")) - 1L) %% every == 0L
  br <- sq[keep]
  lb <- format(d[keep], "%b")
  lb[br < 0] <- paste0(lb[br < 0], "*")
  list(breaks = br, labels = lb)
}

#' p -> significance stars
#' @noRd
.sig_stars <- function(p) {
  ifelse(is.na(p), "",
  ifelse(p < 0.001, "***",
  ifelse(p < 0.01,  "**",
  ifelse(p < 0.05,  "*", ""))))
}

#' Critical |r| for a two-sided test
#' @noRd
.r_crit <- function(n, alpha = 0.05) {
  df <- n - 2
  t <- stats::qt(1 - alpha / 2, df)
  t / sqrt(t^2 + df)
}

#' Window-end offset -> readable date label
#' @noRd
.offset_label <- function(offset, ref_year = 2003) {
  d <- as.Date(paste0(ref_year, "-01-01")) + offset
  paste0(format(d, "%d %b"), ifelse(offset < 0, "*", ""))
}

#' Describe a window in words, e.g. "60 d ending 12 Jul"
window_label <- function(win_len, end_offset, ref_year = 2003) {
  sprintf("%d d ending %s", win_len, .offset_label(end_offset, ref_year))
}

#' "69" or "54-69", for figure subtitles
#' @noRd
.range_label <- function(x) {
  r <- range(x, na.rm = TRUE)
  if (r[1] == r[2]) as.character(r[1]) else paste(r, collapse = "-")
}

#' Keep only the facets in which a binary field actually has a boundary
#'
#' geom_contour() warns ("Zero contours were generated") for any panel where
#' the field is constant - i.e. nothing significant, or everything. Dropping
#' those panels from the contour layer keeps the console clean.
#' @noRd
.contourable <- function(d, col, alpha, facets) {
  d %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(facets))) %>%
    dplyr::filter(any(.data[[col]] < alpha, na.rm = TRUE),
                  any(.data[[col]] >= alpha, na.rm = TRUE)) %>%
    dplyr::ungroup()
}

#' Common x-axis theming for the window-end axis
#' @noRd
.theme_offset_axis <- function() {
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 6.2)
  )
}

#' FIGURE 1 - climate response surface
#'
#' The main figure: correlation as a function of window end day (x) and window
#' length (y), one panel per climate variable x chronology group. Thin contour
#' = p < alpha, thick contour = FDR-adjusted p < alpha, diamond = strongest
#' window in the panel.
#'
#' @param res Output of `moving_window_cor()`.
#' @param chron_type Which chronology, e.g. "BBI".
#' @param mode "raw" or "detrended".
#' @param r_limit Symmetric colour-scale limit. Defaults to the max |r| in the
#'   plotted subset; pass `max(abs(res$r))` to keep raw and detrended
#'   surfaces on one comparable scale.
#' @param label_every Label every n-th month on the x axis.
plot_response_surface <- function(res, chron_type, mode,
                                  r_limit = NULL, alpha = 0.05,
                                  rows = "Climate", cols = "Group",
                                  label_every = 2, ref_year = 2003) {

  d <- res %>%
    dplyr::filter(.data$Chron == chron_type, .data$Mode == mode)
  if (!nrow(d)) stop("No rows for Chron = ", chron_type, ", Mode = ", mode)

  # geom_raster assumes an evenly spaced grid, which is what the default
  # window_lengths / end_offsets give.
  if (length(unique(diff(sort(unique(d$WinLen))))) > 1 ||
      length(unique(diff(sort(unique(d$EndOffset))))) > 1) {
    warning("Uneven window grid: the raster cells will be misaligned.")
  }

  facets <- c(rows, cols)
  bw <- best_windows(d, by = facets)
  ax <- .offset_axis(d$EndOffset, ref_year, every = label_every)
  lim <- if (is.null(r_limit)) max(abs(d$r), na.rm = TRUE) else r_limit

  p <- ggplot2::ggplot(d, ggplot2::aes(.data$EndOffset, .data$WinLen)) +
    ggplot2::geom_raster(ggplot2::aes(fill = .data$r))

  d_p <- .contourable(d, "p", alpha, facets)
  if (nrow(d_p)) {
    p <- p + ggplot2::geom_contour(
      data = d_p, ggplot2::aes(z = as.numeric(.data$p < alpha)),
      breaks = 0.5, colour = "grey25", linewidth = 0.2, na.rm = TRUE)
  }
  d_q <- .contourable(d, "p_fdr", alpha, facets)
  if (nrow(d_q)) {
    p <- p + ggplot2::geom_contour(
      data = d_q, ggplot2::aes(z = as.numeric(.data$p_fdr < alpha)),
      breaks = 0.5, colour = "black", linewidth = 0.55, na.rm = TRUE)
  }

  p +
    ggplot2::geom_vline(xintercept = 0, colour = "grey20",
                        linetype = "22", linewidth = 0.3) +
    ggplot2::geom_point(data = bw, shape = 23, size = 1.6, stroke = 0.5,
                        fill = NA, colour = "black") +
    ggplot2::facet_grid(
      stats::as.formula(paste(rows, "~", cols)),
      labeller = ggplot2::labeller(Climate = CLIM_LABELS)
    ) +
    ggplot2::scale_fill_gradient2(
      low = "#2166AC", mid = "#F7F7F7", high = "#B2182B", midpoint = 0,
      limits = c(-lim, lim), oob = scales::squish,
      name = "Pearson r"
    ) +
    ggplot2::scale_x_continuous(breaks = ax$breaks, labels = ax$labels,
                                expand = c(0, 0)) +
    ggplot2::scale_y_continuous(breaks = pretty(d$WinLen), expand = c(0, 0)) +
    ggplot2::labs(
      x = "Window end (* = previous year)",
      y = "Window length (days)",
      title = paste0(chron_type, " vs ", mode, " climate"),
      subtitle = sprintf("moving windows of %d-%d days; %s overlapping years",
                         min(d$WinLen), max(d$WinLen), .range_label(d$n)),
      caption = paste("dashed line = 1 January of the growth year;",
                      "thin contour p <", alpha,
                      "| thick contour FDR <", alpha,
                      "| diamond = strongest window")
    ) +
    ggplot2::theme_bw(base_size = 9) +
    .theme_offset_axis() +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      panel.spacing = grid::unit(0.4, "lines"),
      legend.key.height = grid::unit(1.2, "lines"),
      strip.background = ggplot2::element_rect(fill = "grey92", colour = NA),
      plot.caption = ggplot2::element_text(colour = "grey35", size = 6.5)
    )
}


#' Extract the yearly climate series of the strongest windows
#'
#' @return Long tibble: one row per (group, chronology, variable, mode, year)
#'   with the chronology index and the climate value of that panel's best
#'   window.
best_window_data <- function(res, daily, chron,
                             chron_cols = c(BBI = "BBI_chron"),
                             clim_vars = c(P = "sum", Tmean = "mean",
                                           Tmax = "mean", Tmin = "mean",
                                           D = "sum"),
                             group_col = "Species", year_col = "Year",
                             nyrs = 30, detrend_method = "difference") {

  chron <- as.data.frame(chron)
  years <- sort(unique(chron[[year_col]]))
  bw <- best_windows(res)

  dplyr::bind_rows(lapply(seq_len(nrow(bw)), function(i) {
    b <- bw[i, ]
    v <- as.character(b$Climate)
    s <- window_series(daily, v, clim_vars[[v]], years,
                       b$EndOffset, b$WinLen,
                       detrend = b$Mode == "detrended",
                       nyrs = nyrs, detrend_method = detrend_method)
    sel <- as.character(chron[[group_col]]) == b$Group
    ch <- tibble::tibble(Year = chron[[year_col]][sel],
                         Index = chron[[chron_cols[[b$Chron]]]][sel])
    dplyr::inner_join(s, ch, by = "Year") %>%
      dplyr::filter(!is.na(.data$Value), !is.na(.data$Index)) %>%
      dplyr::mutate(Group = b$Group, Chron = b$Chron, Climate = b$Climate,
                    Mode = b$Mode, WinLen = b$WinLen,
                    EndOffset = b$EndOffset, r = b$r, p_fdr = b$p_fdr,
                    Window = window_label(b$WinLen, b$EndOffset))
  }))
}



#' FIGURE 2 - chronology and best climate window through time (z-scores)
#'
#' @param flip_negative Multiply the climate z-score by the sign of r, so a
#'   negative relationship still shows as two curves tracking each other. The
#'   panel label says "inverted" where this happened.
plot_best_timeseries <- function(bwdata, chron_type, mode,
                                 flip_negative = TRUE) {

  d <- bwdata %>%
    dplyr::filter(.data$Chron == chron_type, .data$Mode == mode) %>%
    dplyr::group_by(.data$Group, .data$Climate) %>%
    dplyr::mutate(
      flip = flip_negative && .data$r[1] < 0,
      z_clim  = as.numeric(scale(.data$Value)) * ifelse(.data$flip, -1, 1),
      z_chron = as.numeric(scale(.data$Index)),
      lab = sprintf("%s | r = %.2f%s%s", .data$Window, .data$r,
                    .sig_stars(.data$p_fdr),
                    ifelse(.data$flip, " (climate inverted)", ""))
    ) %>%
    dplyr::ungroup()

  labs_df <- d %>%
    dplyr::group_by(.data$Group, .data$Climate) %>%
    dplyr::summarise(lab = .data$lab[1], x = min(.data$Year), .groups = "drop")

  ggplot2::ggplot(d, ggplot2::aes(.data$Year)) +
    ggplot2::geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
    ggplot2::geom_line(ggplot2::aes(y = .data$z_chron, colour = "Chronology"),
                       linewidth = 0.45) +
    ggplot2::geom_line(ggplot2::aes(y = .data$z_clim, colour = "Climate"),
                       linewidth = 0.45) +
    ggplot2::geom_text(data = labs_df,
                       ggplot2::aes(x = .data$x, y = Inf, label = .data$lab),
                       hjust = 0, vjust = 1.4, size = 2, colour = "grey25") +
    ggplot2::facet_grid(Climate ~ Group,
                        labeller = ggplot2::labeller(Climate = CLIM_LABELS)) +
    ggplot2::scale_colour_manual(
      values = c(Chronology = "#1B7837", Climate = "#762A83"), name = NULL) +
    ggplot2::scale_y_continuous(
      expand = ggplot2::expansion(mult = c(0.05, 0.22))) +
    ggplot2::labs(
      y = "z-score",
      title = paste0(chron_type, " and ", mode,
                     " climate of the strongest window, through time")
    ) +
    ggplot2::theme_bw(base_size = 9) +
    ggplot2::theme(
      legend.position = "bottom",
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "grey92", colour = NA)
    )
}


```

# Reading each site's ERA5 file

Two helpers that `CodeMain.qmd` did not need, because it read one hard-coded
file and hard-coded the one incomplete day at the start of it.

```{r era5-readers}

#' Read one ERA5 hourly csv, and recover its grid-cell latitude
#'
#' `CodeMain.qmd` dropped `latitude`/`longitude` immediately and set the
#' latitude by hand. With six sites the file's own latitude is both safer and
#' less work, so it is captured before the column is dropped and returned as
#' an attribute.
#'
#' `read_csv` rather than `read.csv`: these are ~750 000 hourly rows each and
#' there are six of them.
era5_read <- function(path) {

  clim <- readr::read_csv(path, show_col_types = FALSE, progress = FALSE)

  if (!"valid_time" %in% names(clim)) {
    stop("No `valid_time` column in ", path, ". Columns: ",
         paste(names(clim), collapse = ", "))
  }

  # Some CDS exports write valid_time as seconds since the epoch rather than
  # as a timestamp; era5_to_daily() would parse that as NA.
  if (is.numeric(clim$valid_time)) {
    clim$valid_time <- as.POSIXct(clim$valid_time, origin = "1970-01-01",
                                  tz = "UTC")
  }

  lat_file <- if ("latitude" %in% names(clim)) {
    mean(as.numeric(clim$latitude), na.rm = TRUE)
  } else {
    NA_real_
  }

  clim <- clim %>%
    select(-any_of(c("latitude", "longitude", "number", "expver")))

  attr(clim, "lat_file") <- lat_file
  clim
}


#' Trim incomplete days from the START and END of an hourly record only
#'
#' The first day of a CDS request is routinely short an hour, or carries no
#' dewpoint (this is what CodeMain's `>= 1940-01-02 00:00:00` filter was for),
#' and the last day is usually partial. Both would become NA days inside
#' `era5_to_daily(require_full_day = TRUE)` and invalidate every window
#' touching them, which is harmless but noisy.
#'
#' INTERIOR incomplete days are deliberately left in place: `era5_to_daily()`
#' masks them to NA and `window_matrix()` then refuses any window containing
#' them. Dropping their rows here instead would leave a hole in the calendar
#' that the whole offset arithmetic assumes is gap-free, and every window after
#' the hole would silently shift.
#'
#' This cannot make the record entirely NA-free, and is not meant to. With
#' `tp_hour_ending = TRUE` the accumulated fields are shifted back one hour, so
#' `era5_to_daily()` will always report exactly two incomplete days: the
#' calendar day BEFORE the first one (which receives a single accumulated hour
#' and no temperature at all) and the last one (23 accumulated hours, because
#' the 00:00 stamp that closes it belongs to the next request). Both sit
#' outside any year that has a chronology, and every window touching them is
#' refused by `window_matrix()`. The "2 day(s) incomplete" message is expected.
#'
#' @return `clim` with leading/trailing partial days removed.
era5_trim_partial_edges <- function(clim) {

  dt <- clim$valid_time
  if (!inherits(dt, "POSIXct")) dt <- as.POSIXct(as.character(dt), tz = "UTC")
  day <- as.Date(dt, tz = "UTC")

  fields <- intersect(c("t2m", "tp", "d2m", "u10", "v10", "ssrd", "sp"),
                      names(clim))

  n_hours <- tapply(rep(1L, length(day)), day, sum)
  has_all <- Reduce(`&`, lapply(fields, function(f)
    tapply(!is.na(clim[[f]]), day, any)))

  good  <- (n_hours == 24L) & has_all
  dates <- as.Date(names(good))

  if (!any(good)) {
    stop("era5_trim_partial_edges(): no complete day in the record.")
  }

  lo <- dates[which(good)[1]]
  hi <- dates[rev(which(good))[1]]

  dropped <- sum(day < lo | day > hi)
  if (dropped) {
    message("era5_trim_partial_edges(): dropped ", dropped,
            " hour(s) outside ", lo, " .. ", hi, ".")
  }

  clim[day >= lo & day <= hi, , drop = FALSE]
}

```

# Daily climate per site

```{r climate-daily}

# verify the PET code against the published worked example before using it
fao56_example18()

DailyClimBySite <- vector("list", nrow(SITES))
names(DailyClimBySite) <- SITES$Site

for (i in seq_len(nrow(SITES))) {

  s <- SITES[i, ]
  message("\n=== ", s$Site, " (", s$Region, ") <- ", s$ClimPath)

  clim <- era5_read(s$ClimPath)

  # The ERA5 cell's own latitude if the file still carries it, otherwise the
  # mean latitude of that site's trees. Ra changes by roughly 10 % per 5
  # degrees, and this set spans 13.5 degrees, so this must not be shared.
  lat <- attr(clim, "lat_file")
  if (is.finite(lat)) {
    message(sprintf("  ERA5 grid cell at %.4f N (mean tree latitude %.4f N)",
                    lat, s$Lat))
  } else {
    lat <- s$Lat
    message(sprintf("  no latitude column; using the mean tree latitude %.4f N",
                    lat))
  }

  clim <- era5_trim_partial_edges(clim)

  DailyClimBySite[[s$Site]] <- era5_to_daily(clim, tp_hour_ending = TRUE) %>%
    add_water_balance(lat = lat, elev = s$Elev)

  rm(clim)
  invisible(gc())
}

# record coverage per site
purrr::imap(DailyClimBySite, ~ range(.x$Date))

# keep the daily series for the sections that will follow (the drought
# progression and extreme-year work all read `D`), so they need not re-read
# six hourly csvs
dir.create("Output/ClimateData", showWarnings = FALSE, recursive = TRUE)
saveRDS(DailyClimBySite, "Output/ClimateData/DailyClim_AllSites.rds")

```

```{r climate-sanity}

# SANITY CHECK, per site. A PET near 7000 means temperatures were left in
# Kelvin; a PET near 60 means the radiation unit conversion is wrong. For this
# latitude span expect annual PET of roughly 400 (Trondheim) to 750 mm
# (Amiens), annual P of roughly 500-900 mm, and a positive annual D at the
# maritime sites. Katowice and Stockholm can run a yearly deficit; that is
# real, not a bug.
ClimAnnual <- bind_rows(lapply(names(DailyClimBySite), function(site) {
  DailyClimBySite[[site]] %>%
    mutate(Year = as.integer(format(Date, "%Y"))) %>%
    filter(Year > min(Year), Year < max(Year)) %>%   # drop partial end years
    group_by(Year) %>%
    summarise(PET_mm = sum(PET), P_mm = sum(P), D_mm = sum(D), .groups = "drop") %>%
    mutate(Site = site, .before = 1)
}))

ClimAnnual %>%
  group_by(Site) %>%
  summarise(across(c(PET_mm, P_mm, D_mm),
                   list(mean = ~ round(mean(.x, na.rm = TRUE)),
                        min  = ~ round(min(.x,  na.rm = TRUE)),
                        max  = ~ round(max(.x,  na.rm = TRUE)))),
            n_years = sum(!is.na(D_mm)),
            .groups = "drop") %>%
  print(width = Inf)

# and the seasonal cycle of PET, which should peak near 4 mm/day in June at
# Ghent and lower at the northern sites
bind_rows(lapply(names(DailyClimBySite), function(site) {
  DailyClimBySite[[site]] %>%
    mutate(Month = as.integer(format(Date, "%m"))) %>%
    group_by(Month) %>%
    summarise(PET_mm_day = mean(PET, na.rm = TRUE), .groups = "drop") %>%
    mutate(Site = site)
})) %>%
  mutate(Site = factor(Site, levels = SITES$Site)) %>%
  ggplot(aes(Month, PET_mm_day, colour = Site)) +
  geom_line(linewidth = 0.6) +
  scale_x_continuous(breaks = 1:12) +
  labs(y = "PET (mm/day)", title = "Seasonal cycle of Penman-Monteith PET") +
  theme_bw()

```

# Moving-window correlations, per site

```{r climate-correlations}

#### settings ####

CLIM_VARS   <- c(P = "sum", Tmean = "mean", Tmax = "mean", Tmin = "mean",
                 D = "sum")   # D = daily water balance, P - PET
CHRON_COLS  <- c(BBI = "BBI_chron")

WIN_LENGTHS <- seq(20, 120, by = 10)   # window lengths in days

# window end days, as day offsets from 1 January of the growth year:
#   -245 = ~1 May of the PREVIOUS year      0 = 1 January
#    334 = ~30 November of the growth year
# (offsets rather than dates, so a window means the same thing in every year;
#  in leap years it falls one calendar day earlier after 29 February)
END_OFFSETS <- seq(-365, 364)

NYRS_SPLINE <- 30                     # same stiffness as the tree-ring spline
MIN_N       <- 20                     # minimum overlapping years per window

#### run, one site at a time ####

# Per site this is 5 variables x length(WIN_LENGTHS) x length(END_OFFSETS)
# x n_species x 1 chronology (BBI) x 2 modes -- a few minutes, nearly all of
# it in the splines -- and there are six sites. The result of each site is
# written to Output/ClimateCorrelation/<SITE>/ClimCor_<SITE>.rds and reused on
# the next render unless RECOMPUTE is TRUE, so an edit to the figures below
# does not cost another half hour.

ClimCorBySite <- list()

for (site in SITES$Site) {

  chron_site <- ChronsAll %>% filter(as.character(Site) == site)

  if (!nrow(chron_site)) {
    message(site, ": no chronology passed the replication floor -- skipped.")
    next
  }

  siteDir <- file.path(outRoot, site)
  dir.create(siteDir, showWarnings = FALSE, recursive = TRUE)
  cachePath <- file.path(siteDir, paste0("ClimCor_", site, ".rds"))

  if (!RECOMPUTE && file.exists(cachePath)) {
    message(site, ": reusing ", cachePath,
            " (set RECOMPUTE <- TRUE to recompute)")
    ClimCorBySite[[site]] <- readRDS(cachePath)
    next
  }

  message("\n=== ", site, ": ", n_distinct(chron_site$Species),
          " chronolog", if (n_distinct(chron_site$Species) == 1) "y" else "ies",
          ", ", min(chron_site$Year), "-", max(chron_site$Year))

  print(system.time(
    res <- moving_window_cor(
      chron       = chron_site,
      daily       = DailyClimBySite[[site]],
      chron_cols  = CHRON_COLS,
      group_col   = "Species",
      year_col    = "Year",
      clim_vars   = CLIM_VARS,
      win_lengths = WIN_LENGTHS,
      end_offsets = END_OFFSETS,
      nyrs        = NYRS_SPLINE,
      detrend_method = "difference",
      min_n       = MIN_N
    )
  ))

  res <- res %>% mutate(Site = site, .before = 1)
  saveRDS(res, cachePath)
  ClimCorBySite[[site]] <- res
}

ClimCor <- bind_rows(ClimCorBySite) %>%
  mutate(Site = factor(Site, levels = SITES$Site))

saveRDS(ClimCor, file.path(outRoot, "ClimCor_AllSites.rds"))

nrow(ClimCor)

# overlapping years per window, per site (climate starts in 1940, so this is
# really a statement about how far back each chronology reaches)
ClimCor %>%
  group_by(Site) %>%
  summarise(n_min = min(n), n_max = max(n), windows = n(), .groups = "drop")

# how much of each surface is significant, before and after FDR
ClimCor %>%
  group_by(Site, Chron, Mode, Climate, Group) %>%
  summarise(
    n_windows = n(),
    max_r     = round(r[which.max(abs(r))], 3),
    pct_p05   = round(100 * mean(p < 0.05), 1),
    pct_fdr05 = round(100 * mean(p_fdr < 0.05), 1),
    .groups   = "drop"
  ) %>%
  arrange(Site, Chron, Mode, Climate, Group) %>%
  print(n = 400)

# strongest window per site / species / chronology / variable / mode
BestWindows <- best_windows(
  ClimCor, by = c("Site", "Group", "Chron", "Climate", "Mode")
) %>%
  mutate(Window = window_label(WinLen, EndOffset))

BestWindows %>%
  select(Site, Group, Chron, Climate, Mode, Window, WinLen, EndOffset,
         r, n, p, p_fdr) %>%
  print(n = 400)

write_csv(BestWindows, file.path(outRoot, "BestWindows_AllSites.csv"))
write_xlsx(BestWindows, file.path(outRoot, "BestWindows_AllSites.xlsx"))

```

# Figures, per site

```{r climate-correlations-figures, fig.width=11, fig.height=10}

# One colour scale for every surface in the document, so raw and detrended --
# and, more to the point, one site and the next -- are directly comparable.
R_LIMIT <- max(abs(ClimCor$r), na.rm = TRUE)
R_LIMIT

for (site in names(ClimCorBySite)) {

  d       <- ClimCorBySite[[site]]
  siteDir <- file.path(outRoot, site)

  BestData <- best_window_data(
    d,
    DailyClimBySite[[site]],
    ChronsAll %>% filter(as.character(Site) == site),
    chron_cols = CHRON_COLS, clim_vars = CLIM_VARS,
    group_col = "Species", year_col = "Year",
    nyrs = NYRS_SPLINE, detrend_method = "difference"
  )

  saveRDS(BestData, file.path(siteDir, paste0("BestData_", site, ".rds")))

  for (ch in names(CHRON_COLS)) {
    for (md in c("raw", "detrended")) {

      # FIGURE 1 - the response surface (the main figure). The site is added
      # to the title rather than to the facets: with five climate variables
      # down and three species across, a site facet would not fit.
      p_surface <- plot_response_surface(d, ch, md, r_limit = R_LIMIT) +
        labs(title = paste0(site, " - ", ch, " vs ", md, " climate"))

      ggsave(file.path(siteDir, sprintf("1_surface_%s_%s_%s.png", site, ch, md)),
             p_surface, width = 11, height = 10, dpi = 600)

      # FIGURE 2 - chronology and best window through time
      p_ts <- plot_best_timeseries(BestData, ch, md) +
        labs(title = paste0(site, " - ", ch, " and ", md,
                            " climate of the strongest window, through time"))

      ggsave(file.path(siteDir, sprintf("2_timeseries_%s_%s_%s.png", site, ch, md)),
             p_ts, width = 11, height = 8, dpi = 600)
    }
  }

  # show the detrended BBI surface of each site inline
  print(
    plot_response_surface(d, "BBI", "detrended", r_limit = R_LIMIT) +
      labs(title = paste0(site, " - BBI vs detrended climate"))
  )
}

```

# Cross-site overview

The per-site surfaces answer "what is this site sensitive to". Laying one
climate variable out with the sites as rows answers the question the six-site
design was built for: whether the sensitive window is the same one everywhere,
or whether it moves with latitude.

```{r cross-site-surfaces, fig.width=11, fig.height=12}

#' One climate variable, sites down and species across
#'
#' `plot_response_surface()` hard-codes `facet_grid(Climate ~ Group)` and a
#' `labeller` for `Climate`. Rather than fork a 60-line plotting function for
#' one figure, the facet specification is replaced afterwards -- adding a new
#' `facet_grid()` to a ggplot overrides the previous one, labeller included.
plot_surface_bysite <- function(res, chron_type, mode, climate,
                                r_limit = NULL, label_every = 2) {

  d <- res %>% dplyr::filter(.data$Climate == climate)
  if (!nrow(d)) stop("No rows for Climate = ", climate)

  plot_response_surface(d, chron_type, mode, r_limit = r_limit,
                        rows = "Site", cols = "Group",
                        label_every = label_every) +
    ggplot2::facet_grid(Site ~ Group) +
    ggplot2::labs(
      title = paste0(CLIM_LABELS[[climate]], " - ", chron_type, " vs ",
                     mode, " climate"),
      subtitle = sprintf("rows = site, columns = species; %d-%d day windows; %s overlapping years",
                         min(d$WinLen), max(d$WinLen), .range_label(d$n))
    )
}

for (ch in names(CHRON_COLS)) {
  for (md in c("raw", "detrended")) {
    for (v in names(CLIM_VARS)) {

      p <- plot_surface_bysite(ClimCor, ch, md, v, r_limit = R_LIMIT)

      ggsave(file.path(outRoot, sprintf("3_bysite_%s_%s_%s.png", v, ch, md)),
             p, width = 11, height = 12, dpi = 600)
    }
  }
}

# the two that matter most, inline
plot_surface_bysite(ClimCor, "BBI", "detrended", "P", r_limit = R_LIMIT)
plot_surface_bysite(ClimCor, "BBI", "detrended", "D", r_limit = R_LIMIT)

```

```{r cross-site-best, fig.width=12, fig.height=7}

# WHERE the strongest window sits, and how wide it is: the segment runs from
# StartOffset to EndOffset, so a window that moves with latitude shows up as a
# staircase down the panel and a window that does not shows up as a stack.
# Colour is r (same scale as the surfaces) and the number is r again, because
# a colour cannot be read off to two decimals.
best_d <- BestWindows %>%
  filter(Chron == "BBI", Mode == "detrended") %>%
  mutate(Site = factor(Site, levels = rev(SITES$Site)))

ax_best <- .offset_axis(c(best_d$StartOffset, best_d$EndOffset), every = 2)

p_best <- ggplot(best_d, aes(y = Site, colour = r)) +
  geom_vline(xintercept = 0, colour = "grey20", linetype = "22",
             linewidth = 0.3) +
  geom_segment(aes(x = StartOffset, xend = EndOffset, yend = Site),
               linewidth = 1.1) +
  geom_point(aes(x = EndOffset), size = 1.2) +
  geom_text(aes(x = EndOffset, label = sprintf("%.2f", r)),
            hjust = -0.35, size = 1.9, colour = "grey25") +
  # `moving_window_cor()` returns the grouping column as `Group`, not by its
  # original name
  facet_grid(Group ~ Climate,
             labeller = labeller(Climate = CLIM_LABELS)) +
  scale_colour_gradient2(low = "#2166AC", mid = "#F7F7F7", high = "#B2182B",
                         midpoint = 0, limits = c(-R_LIMIT, R_LIMIT),
                         oob = scales::squish, name = "Pearson r") +
  scale_x_continuous(breaks = ax_best$breaks, labels = ax_best$labels,
                     expand = expansion(mult = c(0.04, 0.16))) +
  labs(x = "Window (* = previous year)", y = NULL,
       title = "Strongest detrended-climate window per site and species (BBI)",
       caption = paste("segment = the window itself, from its start to its end;",
                       "dashed line = 1 January of the growth year")) +
  theme_bw(base_size = 9) +
  .theme_offset_axis() +
  theme(panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey92", colour = NA),
        plot.caption = element_text(colour = "grey35", size = 6.5))

p_best
ggsave(file.path(outRoot, "4_best_windows_bysite.png"), p_best,
       width = 12, height = 7, dpi = 600)

```

# What this wrote

```
Output/ClimateData/DailyClim_AllSites.rds     daily climate + PET + D, per site
Output/ClimateCorrelation/
  0_chronologies_allsites.png                 every BBI chronology, site x species
  0_replication_allsites.png                  and its replication
  Chrons_AllSites.{csv,rds}                   the chronologies themselves
  ClimCor_AllSites.rds                        every correlation field, all sites
  BestWindows_AllSites.{csv,xlsx}             strongest window per panel
  3_bysite_<VAR>_<CHRON>_<MODE>.png           one variable, sites as rows
  4_best_windows_bysite.png                   strongest r per site and species
  <SITE>/
    ClimCor_<SITE>.rds                        that site's field (the cache)
    BestData_<SITE>.rds                       yearly series of its best windows
    1_surface_<SITE>_<CHRON>_<MODE>.png       the response surface
    2_timeseries_<SITE>_<CHRON>_<MODE>.png    chronology vs best window
```

`ClimCor_AllSites.rds` is the input the sections after this one in
`CodeMain.qmd` will want: it has the same columns as CodeMain's `ClimCor` plus
a leading `Site`.
